\usemodule[simplefonts][size=12pt]
\setmainfont[Crimson]
\setuppapersize[letter][letter]
\setuplayout[width=6.5in,height=10in,topspace=0.5in,backspace=1in,
  header=0.5in,footer=0.5in,bottomspace=0.5in]
\setuppagenumbering[location=]
\setuphead[section][number=no,style=\tfb,after={\blank[0.25cm]}]
\setuphead[subsection][style=\tfa,after={\blank[0.25cm]}]

\starttext

\section{Detailed Overview of Tasks}

\subsection{Research}

\startitemize[1]
\item Model \& Model Background: Determining which model to study and
  understanding the physical basis and real-world applications and
  implications of the model.
\item Math Background: Understanding the theory of dynamical systems and
  bifurcation theory, particularly as applied to modeling.
\item Qualified Scientist: Contacting Dr.\ Baer in order to consult on
  mathematical and computational aspects of the model and possibly another
  scientist qualified in the field of the model to consult on the
  implications of the model.
\stopitemize

\subsection{Design}

\startitemize[1]
\item Question \& Hypothesis: Based on background information, predicting a
  possible real-world implication of the analysis of the model with a slow
  variable
\item Procedural Design: Setting up the programs needed to perform
  calculations and analyze the output.
\stopitemize

\subsection{Experimentation}

\startitemize[1]
\item Experiment \& Computation: Writing any necessary code and performing
  necessary calculations.
\item Analysis: Determining what the results imply. May involve writing code
  to help visualize or analyze the results and consultation with a qualified
  scientist about the most likely physically possible scenarios.
\stopitemize

\subsection{Report \& Poster}

\startitemize[1]
\item Report: Writing up a detailed research report, with an emphasis on
  real-world applications and clear explanations of the mathematics and
  computer science involved.
\item Poster: Will be based on report.
\item Presentation: Both short and long versions; emphasis on explaining
  simply and concisely the details of the experiment and background.
\stopitemize

\stoptext